Emergence of the Southeast Asian islands as a driver for Neogene cooling

### 2020-04-13 Manuscript/Manuscript.tex

% Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.}

@@

% Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
\begin{abstract}
The steep topography, tropical climate, and mafic lithologies of the Southeast Asian islands contribute to efficient chemical weathering and carbon sequestration. Concurrent with global cooling since the mid-Miocene, ongoing collision between the Sunda-Banda arc and Australia has increased the area of subaerially-exposed land in the region. Here we integrate lithology into a coupled weathering-climate model to assess how the global climate system responds to changes in paleogeography. We then use paleoshoreline data as boundary conditions to quantify the change in steady-state \pCOtwo associated with the tectonic emergence of the Southeast Asian islands since the mid-Miocene. We find that the modeled decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain long-term climatic cooling over the Neogene.
\end{abstract}

@@

\dropcap{T}he Southeast Asian islands (SEAI) have an out-sized contribution to modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Zachos2008a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic outgassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1992a, Galy2007a}. Here we instead propose that SEAI emergence was itself sufficient to explain long-term climatic cooling over the Neogene.

Over geologic time-scales, \COtwo enters Earth's ocean--atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, with a smaller sink through organic carbon burial \cite{Kump1997a}. Chemical weathering delivers alkalinity and cations to the ocean which drives carbon sequestration through carbonate precipitation. Steady-state \pCOtwo is set at the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo \cite{Kump1997a}. Conversely, as regions with high carbon sequestration potential emerge on Earth's surface, \pCOtwo decreases resulting in other weathering sinks to diminish leading to a new lower steady-state \pCOtwo.

Topography, climate, and lithology all effect chemical weathering. High-relief regions lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more sensitive to climate change \cite{Gabet2009a, West2012a}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/regolith_schematic.pdf}
    \caption{A schematic representation of the silicate weathering component of GEOCLIM in a single profile at steady-state. A rock particle leaves the unweathered bedrock with production rate $P_{r}$, and transits vertically through a regolith of height $h$. Regolith production and physical erosion ($E_{p}$) are equal at steady-state. As the particle transits upwards, some fraction of the primary phases ($x$) are chemically weathered ($W$) with the flux of Ca and Mg corresponding to that weathering rate multiplied by the concentration of Ca+Mg in unweathered bedrock ($\chi_{CaMg}$). Details of the formulation for the silicate weathering component of GEOCLIM can be found in \MM.}
    \label{fig:regolith_schematic}
\end{figure}

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered. We therefore implement variable bedrock Ca+Mg concentration into GEOCLIM (\SI). The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) \cite{Hartmann2012a} and is represented by 6 categories: metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment. Each land pixel is assigned these lithologic categories at a resolution of 0.1\degrees $\times$ 0.1\degrees. The Ca+Mg concentrations of felsic, intermediate, and mafic lithologies are assigned based on the mean of data compiled from EarthChem (\url{www.earthchem.org/portal}). Given that GLiM does not distinguish ultramafic lithologies, such rocks are grouped with mafic rocks. As a result, the Ca+Mg concentration is likely an underestimate in regions of obducted ophiolites, such that the estimated effect of these regions on changing steady-state \pCOtwo is conservative. The weathering of carbonate does not contribute to long-term \COtwo consumption and its Ca+Mg concentration is ignored. The Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies are more difficult to define, since their chemical composition is strongly dependent on protolith composition and, in the case of siliciclastic sediment, the degree of previous chemical depletion. We explore a range of feasible Ca+Mg concentrations for metamorphic rocks and siliciclastic sediment during calibration of the silicate weathering component.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic pixels to vary within reasonable bounds (\SI). This results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 5,381 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo outgassing flux as they should with the long-term carbon cycle being at steady-state (\SI).

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=11.4cm]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the Southeast Asian islands (also referred to as the Maritime Continent in climate science literature) from the mid-Miocene to present. Past shorelines are shown in \textit{A} with associated land area summarized in \textit{B}. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in \textit{C}.}
    \label{fig:shoreline_growth}
\end{figure*}

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 5,381 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (\SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. Glaciation thresholds \cite{DeConto2008a} are shown on the x-axis.}
    \label{fig:scenario_pCO2}
\end{figure*}

Using the 5,381 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 378--447~ppm \pCOtwo for 5~Ma, 435--532~ppm for 10~Ma, and 493--709~ppm for 15~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$493--709~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (\SI), indicating that the increase in subaerially-exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

The results of our SEAI scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of low-relief basaltic lavas on steady-state \pCOtwo \cite{Kent2013a}, we replace mafic volcanics associated with the Deccan Traps, Ethiopian Traps, and Columbia River Basalts with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling \cite{Kent2013a}, its contribution is less than that of the wetter and higher-relief SEAI.

Other hypotheses to explain ice sheet growth over the Neogene invoke changes in ocean/atmosphere circulation including: further climatic isolation of Antarctica due to strengthening of the circumpolar current \cite{Shevenell2004a}; increased atmospheric moisture in the Northern Hemisphere due to intensified thermohaline circulation following Panama Isthmus emergence \cite{Haug1998a}; and cooling of North America resulting from a strengthened Walker Circulation associated with SEAI emergence \cite{Molnar2015a}. Such changes in ocean/atmosphere circulation are likely to modulate \pCOtwo thresholds for glacial initiation and ice sheet growth \cite{DeConto2008a}. However, the prolonged timescale of the cooling trend since 15~Ma (Fig. \ref{fig:shoreline_growth}C) is most readily attributable to decreasing \pCOtwo associated with evolving geological sources and sinks of carbon, modulated by the silicate weathering feedback \cite{Kump1997a}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Manuscript/Figures/weatherability_curves.pdf}
    \caption{Weatherability curves for the ``paleo-SEAI'' scenarios shown in Figure \ref{fig:scenario_pCO2}. Details on how these curves were generated are described in \MM. Each of the 4 curves represent a different tectonic boundary condition (i.e. the reconstructed paleoshorelines of the SEAI; Fig. \ref{fig:shoreline_growth}A), and shows the volcanic degassing flux necessary to balance the silicate weathering flux at a given \pCOtwo (and vice versa). Point B represents the pre-Industrial, in which \pCOtwo is 286~ppm. The arrow from Point A$_{1}$ to B represents the ``weatherability only'' scenario, in which global weatherability changes as the SEAI emerge, but the volcanic degassing flux does not change over the past 15~m.y. In this scenario, the \pCOtwo decreases from the value dictated by the 15~Ma ``paleo-SEAI'' weatherability curve ($\sim$570~ppm). The arrow from Point A$_{2}$ to B instead represents the ``degassing only'' scenario, in which global weatherability remains the same as the pre-Industrial, but the same change in \pCOtwo as the ``weatherability only'' scenario is achieved by decreasing the volcanic degassing flux. In this scenario, the volcanic degassing flux decreases from the value dictated by the pre-Industrial weatherability curve ($\sim$13\% greater than the pre-Industrial). The arrow from Point A$_{3}$ to B represents the ``weatherability and degassing'' scenario, in which global weatherability changes as the SEAI emerge and the volcanic degassing flux decreases from a value $\sim$4\% greater than the pre-Industrial. This change in the volcanic degassing flux corresponds to the \pCOtwo threshold estimated for Antarctic glaciation \cite{DeConto2008a} on a planet with our 15~Ma ``paleo-SEAI'' paleoshorelines.}
    \label{fig:weatherability_curves}
\end{figure}

A decrease in volcanic degassing \cite{Berner1983a} has also been proposed as a driver for Neogene cooling. Proxy-based estimates of the evolution of volcanic degassing fluxes throughout the Neogene are inconsistent with each other, such that not even the sign of the change in volcanic degassing over the past $\sim$15~m.y. is without ambiguity \cite{Godderis2017c}. For example, it has been estimated that the volcanic degassing flux was $\sim$25\% lower \cite{Cogne2006a} and $\sim$10\% higher \cite{Van-Der-Meer2014a} at 15~Ma relative to the present day.

Nevertheless, our model framework provides us with the opportunity to estimate the magnitude of volcanic degassing flux decrease necessary to achieve the same change in \pCOtwo predicted for the increase in global weatherability associated with the emergence of the SEAI over the past 15~m.y. If we use the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI), GEOCLIM estimates a volcanic degassing flux for the pre-Industrial of 4.1$\times$10$^{12}$~mol/yr. If we then assume that this volcanic degassing flux did not change over the past 15~m.y., then GEOCLIM estimates that the increase in global weatherability associated with the emergence of the SEAI lead to a change in \pCOtwo of $\sim$280~ppm (``weatherability only'' scenario in Fig. \ref{fig:weatherability_curves}). If we instead assume that global weatherability did not change over the past 15~m.y., then we estimate that the volcanic degassing flux needs to have been $\sim$13\% greater at 15~Ma relative to the pre-Industrial to drive the same $\sim$280~ppm change in \pCOtwo (``degassing only'' scenario in Fig. \ref{fig:weatherability_curves}).

This $\sim$13\% value is higher than 10\%, the highest current estimate for the volcanic degassing flux at 15~Ma relative to the present day \cite{Van-Der-Meer2014a}, suggesting that reasonable changes in the volcanic degassing flux alone are insufficient to drive the majority of cooling observed over the Neogene. However, changes in the volcanic degassing flux would have modulated changes in global weatherability associated with the emergence of the SEAI. In a hypothetical scenario in which \pCOtwo was at the threshold for Antarctic glaciation \cite{DeConto2008a} at 15~Ma and the SEAI are emerging, the volcanic degassing flux would only have had to have been $\sim$4\% higher relative to the pre-Industrial (``weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}).

@@

Himalayan uplift \cite{Raymo1992a, Galy2007a} has also been proposed as drivers for Neogene cooling. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalayan lithologies \cite{Colleps2018a}, but could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in the SEAI during arc-continent collision. Enhanced organic matter burial in the Bay of Bengal due to high sedimentation rates could also have contributed to Neogene cooling \cite{Galy2007a}.

>>>>

% Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality. The primary goal of the significance statement is to explain the relevance of the work in broad context to a broad readership. The significance statement appears in the paper itself and is required for all research papers.
\significancestatement{The Southeast Asian islands are a hotspot of \COtwo consumption via silicate weathering in the present day. Since $\sim$15 million years ago, these islands have been increasing in size at the same time that Earth's climate has been cooling. Here we test the hypothesis that global cooling over the past 15 million years has been driven by tectonic emergence of the Southeast Asian islands. To do so, we improve a silicate weathering model, then couple it to a climate model. We find that the majority of cooling over the past 15 million years can be explained by emergence of the Southeast Asian islands.}

@@

% Please provide an abstract of no more than 250 words in a single paragraph. Abstracts should explain to the general reader the major contributions of the article. References in the abstract must be cited in full within the abstract itself and cited in the text.
\begin{abstract}
Steep topography, a tropical climate, and mafic lithologies contribute to efficient chemical weathering and carbon sequestration in the Southeast Asian islands. Ongoing arc-continent collision between the Sunda-Banda arc and Australia has increased the area of subaerially-exposed land in the region since the mid-Miocene. Concurrently, Earth's climate has been cooling, marked with the onset of Northern Hemisphere glaciation 3 million years ago. Here we integrate lithology into a coupled weathering-climate model to assess how the global climate system responds to changes in weatherability associated with paleogeographic changes. We then use paleoshoreline data as boundary conditions to quantify the change in steady-state \pCOtwo associated with the tectonic emergence of the Southeast Asian islands since the mid-Miocene. We find that the modeled decrease in \pCOtwo due to these paleogeographic changes is sufficient to explain the majority of long-term climatic cooling over the Neogene.
\end{abstract}

@@

\dropcap{T}he Southeast Asian islands (SEAI) have an out-sized contribution to modern chemical weathering fluxes. The confluence of steep topography, a warm and wet tropical climate, and the presence of mafic lithologies results in high fluxes of Ca and Mg cations and associated \COtwo consumption \cite{Gaillardet1999a, Hartmann2009a, Milliman2013a, Hartmann2014a}. There has been a significant increase of subaerially-exposed land area within the region since the Miocene associated with ongoing arc-continent collision between Australia and the Sunda-Banda arc system \cite{Molnar2015a, Hall2017a, Macdonald2019a}. Concurrently, after the Miocene Climatic Optimum, a cooling trend began ca. 15~Ma and accelerated over the past 4 million years (m.y.) culminating with the onset of Northern Hemisphere glaciation \cite{Shackleton1984a, Zachos2001a}. Many hypotheses have been proposed to explain this cooling trend including changes in ocean/atmosphere circulation \cite{Haug1998a, Shevenell2004a, Molnar2015a}, a decrease in volcanic degassing \cite{Berner1983a}, or uplift in the Himalaya \cite{Raymo1988a, Galy2007a}. Here we instead propose that SEAI emergence was itself sufficient to explain long-term climatic cooling over the Neogene.

Over geologic time-scales, \COtwo enters Earth's ocean--atmosphere system primarily via volcanism and metamorphic degassing, and leaves primarily through the chemical weathering of silicate rocks, with a smaller sink through organic carbon burial \cite{Kump2000a}. Chemical weathering delivers alkalinity and cations to the ocean which drives carbon sequestration through carbonate precipitation. Steady-state \pCOtwo is set at the \pCOtwo level at which \COtwo sinks are equal to sources. As \COtwo sinks are removed and \pCOtwo rises, temperature increases and the hydrological cycle is invigorated, causing other weathering sinks to increase until a new steady-state is achieved at higher \pCOtwo \cite{Kump1997a}. Conversely, as regions with high carbon sequestration potential emerge on Earth's surface, \pCOtwo decreases resulting in other weathering sinks to diminish leading to a new lower steady-state \pCOtwo.

Topography, climate, and lithology all effect chemical weathering. High-relief regions lack extensive regolith development, and thus tend to have reaction-limited weathering regimes that are more sensitive to climate change \cite{Gabet2009a, West2012a, Maher2014a}. In warm and wet regions, mineral dissolution kinetics are faster leading to enhanced chemical weathering \cite{Lasaga1994a, West2012a}. Mafic rocks have higher Ca and Mg concentrations and dissolution rates than felsic rocks, and thus the potential to more efficiently sequester carbon through silicate weathering \cite{Dessert2003a}. These factors have lead to the proposal that arc-continent collisions within the tropical rain belt have been important in enhancing global weatherability, lowering atmospheric \pCOtwo, and initiating glacial climate over the past 520~m.y. \cite{Jagoutz2016a, Swanson-Hysell2017a, Macdonald2019a} and perhaps in the Neoproterozoic as well \cite{Park2019b}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/regolith_schematic.pdf}
    \caption{A schematic representation of the silicate weathering component of GEOCLIM in a single profile at steady-state. A rock particle leaves the unweathered bedrock with production rate $P_{r}$, and transits vertically through a regolith of height $h$. Regolith production and physical erosion ($E_{p}$) are equal at steady-state. As the particle transits upwards, some fraction of the primary phases ($x$) are chemically weathered ($W$) with the flux of Ca+Mg corresponding to that weathering rate multiplied by the concentration of Ca+Mg in unweathered bedrock ($\chi_{CaMg}$). Details of the formulation for the silicate weathering component of GEOCLIM can be found in \MM.}
    \label{fig:regolith_schematic}
\end{figure}

The silicate weathering component of GEOCLIM calculates \COtwo consumption resulting from silicate weathering for subaerially-exposed land. We assume that Ca and Mg are the only cations that consume \COtwo over geologic time-scales, such that each mole of Ca or Mg that is dissolved by silicate weathering consumes one mole of \COtwo. In previous versions of the model, silicate weathering was a function of temperature and runoff only, and all bedrock was assigned identical chemical compositions \cite{Godderis2017c}. More recent versions of GEOCLIM implement regolith development and soil shielding (Fig. \ref{fig:regolith_schematic}), which introduces a dependence on erosion (and therefore topographic slope) \cite{Maffre2018a}. While this introduction of regolith development into GEOCLIM is important for assessing the impact of tropical arc-continent collisions on \pCOtwo, the relatively high Ca+Mg concentration in arc rocks relative to other lithologies must also be considered. We therefore implement variable bedrock Ca+Mg concentration into GEOCLIM (\SI). The spatial distribution of lithologies is sourced from the Global Lithologic Map (GLiM) \cite{Hartmann2012a} and is represented by 6 categories: metamorphic, felsic, intermediate, mafic, carbonate, and siliciclastic sediment. Each land pixel is assigned these lithologic categories at a resolution of 0.1\degrees $\times$ 0.1\degrees. The Ca+Mg concentrations of felsic, intermediate, and mafic lithologies are assigned based on the mean of data compiled from EarthChem (\url{www.earthchem.org/portal}). Given that GLiM does not distinguish ultramafic lithologies, such rocks are grouped with mafic rocks. As a result, the Ca+Mg concentration is likely an underestimate in regions of obducted ophiolites, such that the estimated effect of these regions on changing steady-state \pCOtwo is conservative. The weathering of carbonate does not contribute to long-term \COtwo consumption and its Ca+Mg concentration is ignored. The Ca+Mg concentrations of metamorphic and siliciclastic sediment lithologies are more difficult to define, since their chemical composition is strongly dependent on protolith composition and, in the case of siliciclastic sediment, the degree of previous chemical depletion. We explore a range of feasible Ca+Mg concentrations for metamorphic rocks and siliciclastic sediment during calibration of the silicate weathering component.

@@

The values of four parameters within the silicate weathering component that modify the dependence of silicate weathering on temperature, runoff, and regolith thickness are poorly constrained. Rather than prescribing values, we allow these parameters and the Ca+Mg concentration of metamorphic and siliciclastic pixels to vary within reasonable bounds (\SI). This results in 93,600 unique combinations of these six parameters. For each combination, we compute spatially-resolved long-term \COtwo consumption associated with Ca+Mg fluxes using present-day runoff, temperature, and slope. We sum computed \COtwo consumption over watersheds for which data-constrained estimates are available \cite{Gaillardet1999a, Moquet2018a}, then calculate the coefficient of determination ($r^{2}$) between computed and measured \COtwo consumption in each of these watersheds. After eliminating parameter combinations that result in low $r^{2}$, 573 parameter combinations remain. The resulting global \COtwo consumption of these model runs all overlap with independently-derived estimates of global \COtwo degassing flux \cite{Gerlach2011a} as they should with the long-term carbon cycle being at steady-state (\SI).

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=0.60\linewidth]{Figures/shoreline_growth.pdf}
    \caption{The emergence of the Southeast Asian islands (also referred to as the Maritime Continent in climate science literature) from the mid-Miocene to present. Past shorelines are shown in \textit{A} with associated land area summarized in \textit{B}. A significant increase in area over the past 5 million years is coincident with cooling and the onset of Northern Hemisphere glaciation as reflected in the benthic oxygen isotope record \cite{Zachos2008a} shown in \textit{C}.}
    \label{fig:shoreline_growth}
\end{figure*}

@@

\begin{figure*}[h]
    \centering
    \includegraphics[width=1\linewidth]{Figures/scenario_pCO2.pdf}
    \caption{Steady-state \pCOtwo estimates from GEOCLIM for the various scenarios discussed in the text. For each of the seven scenarios, each point represents an estimate from one of the 573 unique parameter combinations that resulted in reasonable total global \COtwo consumption and most closely matched estimates of present-day \COtwo consumption in 80 watersheds around the world (\SI). The box encloses the middle 50\% of the \pCOtwo estimates (i.e. the interquartile range), and the notch represents the median with its 95\% confidence interval. The whiskers extend to the 2.5 and 97.5 percentile values. Glaciation thresholds \cite{DeConto2008a} are shown on the x-axis.}
    \label{fig:scenario_pCO2}
\end{figure*}

Using the 573 unique parameter combinations, the ``paleo-SEAI'' scenarios resulted in 391--434~ppm \pCOtwo for 5~Ma, 457--516~ppm for 10~Ma, and 526--678~ppm for 15~Ma (Fig. \ref{fig:scenario_pCO2}). These results indicate a progressive decrease in \pCOtwo over the Neogene associated with the emergence of the SEAI, and suggest that without this emergence, pre-industrial \pCOtwo would have been $\sim$526--678~ppm. Proxy-based estimates of the magnitude and trajectory of \pCOtwo change from the Miocene to the Pliocene are variable between techniques and associated assumptions underlying their interpretation \cite{Foster2017a}. The modeled \pCOtwo values for 15~Ma resemble the higher end of proxy-based \pCOtwo estimates for the early-mid-Miocene (\SI), indicating that the increase in subaerially-exposed land area and tectonic topography of the SEAI is sufficient to explain long-term cooling of Earth's climate over the Neogene. The \pCOtwo threshold for Antarctic glaciation is estimated to be $\sim$750~ppm with that for Northern Hemisphere glaciation being significantly lower at $\sim$280~ppm \cite{DeConto2008a}. These modeled values of decreasing \pCOtwo are therefore consistent with the record of Neogene climate with Miocene ice sheets on Antarctica \cite{Sugden1995a} followed by Northern Hemisphere ice sheets developing in the Pliocene \cite{Haug2005a} as \pCOtwo subsequently decreased.

The results of our SEAI scenarios highlight the importance of the combination of topography, runoff, and lithology in setting Earth's climate state. To independently explore the effect of the modern-day surface exposure of low-relief basaltic lavas on steady-state \pCOtwo \cite{Kent2013a}, we replace mafic volcanics associated with the Deccan Traps, Ethiopian Traps, and Columbia River Basalts with the Ca+Mg concentration of bulk continental crust in GEOCLIM (Fig. \ref{fig:scenario_pCO2}). The resulting \pCOtwo is $<$500~ppm, indicating that, while the presence of mafic rocks in these igneous provinces modulates steady-state \pCOtwo as has been suggested to be important for Paleogene cooling \cite{Kent2013a}, its contribution is less than that of the SEAI, which have high-relief and a wet tropical climate in conjunction with high Ca+Mg lithologies.

Other hypotheses to explain ice sheet growth over the Neogene invoke changes in ocean/atmosphere circulation including: further climatic isolation of Antarctica due to strengthening of the circumpolar current \cite{Shevenell2004a}; increased atmospheric moisture in the Northern Hemisphere due to intensified thermohaline circulation following Panama Isthmus emergence \cite{Haug1998a}; and cooling of North America resulting from a strengthened Walker Circulation associated with SEAI emergence \cite{Molnar2015a}. Such changes in ocean/atmosphere circulation are likely to modulate \pCOtwo thresholds for glacial initiation and ice sheet growth \cite{DeConto2008a}. However, the prolonged timescale of the cooling trend since 15~Ma (Fig. \ref{fig:shoreline_growth}C) is most readily attributable to decreasing \pCOtwo associated with evolving geological sources and sinks of carbon, modulated by the silicate weathering feedback \cite{Walker1981a, Raymo1991a, Berner1997a, Kump1997a, Berner2001a}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{Manuscript/Figures/weatherability_curves.pdf}
    \caption{Weatherability curves for the ``paleo-SEAI'' scenarios shown in Figure \ref{fig:scenario_pCO2}. Details on how these curves were generated are described in \MM. Each of the 4 curves represent a different tectonic boundary condition (i.e. the reconstructed paleoshorelines of the SEAI; Fig. \ref{fig:shoreline_growth}A), and shows the volcanic degassing flux necessary to balance the silicate weathering flux at a given \pCOtwo (or vice versa). Point B represents the pre-Industrial, in which \pCOtwo is 286~ppm. The arrow from Point A$_{1}$ to B represents the ``weatherability only'' scenario, in which global weatherability changes as the SEAI emerge, but the volcanic degassing flux does not change over the past 15~m.y. In this scenario, the \pCOtwo decreases from the value dictated by the 15~Ma ``paleo-SEAI'' weatherability curve ($\sim$570~ppm). The arrow from Point A$_{2}$ to B instead represents the ``degassing only'' scenario, in which global weatherability remains the same as the pre-Industrial, but the same change in \pCOtwo as the ``weatherability only'' scenario is achieved by decreasing the volcanic degassing flux. In this scenario, the volcanic degassing flux decreases from the value dictated by the pre-Industrial weatherability curve ($\sim$13\% greater than the pre-Industrial). The arrow from Point A$_{3}$ to B represents the ``weatherability and degassing'' scenario, in which global weatherability changes as the SEAI emerge and the volcanic degassing flux decreases from a value $\sim$4\% greater than the pre-Industrial.}
    \label{fig:weatherability_curves}
\end{figure}

A decrease in volcanic degassing \cite{Berner1983a} has also been proposed as a driver for Neogene cooling. However, proxy-based estimates of the evolution of volcanic degassing fluxes throughout the Neogene are inconsistent with each other, such that not even the sign of the change in volcanic degassing over the past $\sim$15~m.y. is without ambiguity \cite{Godderis2017c}. For example, it has both been estimated that the volcanic degassing flux was $\sim$25\% lower \cite{Cogne2006a} and $\sim$10\% higher \cite{Van-Der-Meer2014a} at 15~Ma relative to the present day.

Nevertheless, our model framework provides us with the opportunity to estimate the decrease in volcanic degassing flux necessary to achieve the same change in \pCOtwo predicted for the increase in global weatherability associated with the emergence of the SEAI over the past 15~m.y. If we use the parameter combination that had the highest $r^{2}$ between computed and measured \COtwo consumption in watersheds around the world during calibration (\textit{Calibration} and \SI), GEOCLIM estimates a volcanic degassing flux for the pre-Industrial of 4.1$\times$10$^{12}$~mol/yr. If we then assume that this volcanic degassing flux did not change over the past 15~m.y., then GEOCLIM estimates that the increase in global weatherability associated with the emergence of the SEAI lead to a change in \pCOtwo of $\sim$280~ppm (``weatherability only'' scenario in Fig. \ref{fig:weatherability_curves}). If we instead assume that global weatherability did not change over the past 15~m.y., then we estimate that the volcanic degassing flux needs to have been $\sim$13\% greater at 15~Ma relative to the pre-Industrial to drive the same $\sim$280~ppm change in \pCOtwo (``degassing only'' scenario in Fig. \ref{fig:weatherability_curves}).

This $\sim$13\% value is higher than 10\%, the highest current estimate for the volcanic degassing flux at 15~Ma relative to the present day \cite{Van-Der-Meer2014a}, suggesting that reasonable changes in the volcanic degassing flux alone are insufficient to drive the majority of cooling observed over the Neogene. However, changes in the volcanic degassing flux would have modulated changes in \pCOtwo associated with changes in global weatherability. In a hypothetical scenario in which \pCOtwo was at the threshold for Antarctic glaciation \cite{DeConto2008a} at 15~Ma and the SEAI are emerging, the volcanic degassing flux would only have had to have been $\sim$4\% higher relative to the pre-Industrial to drive \pCOtwo from 750~ppm to 286~ppm (``weatherability and degassing'' scenario in Fig. \ref{fig:weatherability_curves}).

@@

Himalayan uplift \cite{Raymo1992a, Galy2007a} has also been proposed as drivers for Neogene cooling. Rising marine \SrSr has been associated with increasing weathering of the high \SrSr Himalayan lithologies \cite{Raymo1992a}. This increase in \SrSr since ca. 35~Ma is followed by a decrease in slope ca. 15~Ma that has been attributed to exhumation of relatively low \SrSr and high \OsOs Lesser Himalayan lithologies \cite{Quade1997a, Colleps2018a}, but could also be at least partially driven by the exhumation of mafic and organic-rich forearc sediments in the SEAI during arc-continent collision. Enhanced organic matter burial in the Bay of Bengal due to high sedimentation rates could also have contributed to Neogene cooling \cite{Galy2007a}.
